### 2013-01-15 sections/evaluation.tex
re-read the paper; few grammatical corrections

It can be seen that the performance of the filter stage in \texttt{nfql} is
comparable to that of \texttt{flow-tools} and SiLK. SiLK takes less time on
lower ratios, but then SilK and \texttt{nfdump} also use their own file
format. \texttt{nfdump} appears to be significantly faster than the rest. This
is because \texttt{nfdump} uses the \texttt{lzo} compression scheme. Adding
\texttt{lzo} compression will likely improve \texttt{nfql}'s filter
performance. Note that all the tools were single-threaded
in this evaluation, and did not completely utilize the $24$ available cores.

@@

The third set of queries stress the group filter stage. We reuse
the filter and grouper queries that produce a $1.0$ ratio to allow the group
filter to receive the entire trace as input. This means, each flow record of
the original trace now becomes a group record for the group filter.  The group
filter then reuses the same varying values of the \texttt{packet} field offset
to control the amount of groups that are filtered ahead. The filtered groups
are again written as \texttt{flow-tools} files using the same \texttt{zlib}
compression level. The ratio of the number of output filtered groups to the
number of the input group is plotted against time. The results are shown in
Fig. \ref{fig:benchmarks-groupfilter}.

@@

The fourth set of queries stress the merger stage. We reuse the filter,
grouper and group filter queries that produce a $1.0$ ratio. These queries are
then run in two separate branches to produce identical filtered group records.
The merger then applies match rules to produce different output to input
ratios. The groups that are merged are again written as \texttt{flow-tools}
files using the same \texttt{zlib} compression level. The ratio of the number
of merged groups to twice\footnote{Each branch pushes the entire trace as an
input to the merger.} the number of flow records in the original trace is
plotted against time. A data point for SiLK for the $0.2$ ratio is not
available since the \ac{NFQL} query executed at that data point uses
\texttt{OR} expressions, which are not supported by SiLK.

@@

It can be seen that the merger is the most performance critical stage of the
\ac{NFQL} pipeline thus far. It is due to the fact that the merger is working
on twice the number of flow records than any other previous stage. In
addition, each branch is writing the results of the filter, grouper and group
filter stage to \texttt{flow-tools} files. As a result, the amount of disk I/O
involved is twice as much as well. Even though each branch is delegated to a
separate core using affinity masks, most of time is taken in writing these
results to the file. These results although look less promising, they are way
better than the original \texttt{nfql} merger implementation. The optimized
merger implementation takes advantage of sorted nature of filtered groups and
therefore can significantly reduce the number of merger matches. It also
writes a merged group record to a file only once despite the number of times
it has matched. Without these optimizations, running such queries on the
merger kept the CPU churning for days without results.

The last set of queries stress the ungrouper stage. They reuse the entire
merger queries as is, but enable the ungrouper as well. This means, that the
ungrouper now attempts to unfold the merged groups returned by the merger to
write their member flow records to \texttt{flow-tools} files.  However, since
the merger receives each flow record as its own filtered group, each merged
group has only one member. As a result, the ungrouper ends up rewriting the
merged groups as \texttt{flow-tools} files using the same \texttt{zlib}
compression level. This means that the execution engine end up taking twice
the amount of time than the merger. It is important to note, that SiLK does
not have such an equivalent ungrouping tool and is therefore not considered in
this final evaluation.

>>>>

It can be seen that the performance of the filter stage in \texttt{nfql} is
comparable to that of \texttt{flow-tools} and SiLK. SiLK takes less time on
lower ratios, but then SilK and \texttt{nfdump} also use their own file
format. \texttt{nfdump} appears to be significantly faster than the rest. This
is because \texttt{nfdump} uses the \texttt{lzo} compression scheme. It goes
to show that adding \texttt{lzo} compression will likely improve
\texttt{nfql}'s filter performance. Note that all the tools were
single-threaded in this evaluation, and did not completely utilize the $24$
available cores.

@@

The third set of queries stress the group filter stage. We reuse
the filter and grouper queries that produce a $1.0$ ratio to allow the group
filter to receive the entire trace as input. This means, each flow record of
the original trace now becomes a group record for the group filter.  The group
filter then reuses the same varying values of the \texttt{packet} field offset
to control the amount of groups that are filtered ahead. The filtered groups
are again written as \texttt{flow-tools} files using the same \texttt{zlib}
compression level. The ratio of the number of output filtered groups to the
number of the input groups is plotted against time. The results are shown in
Fig. \ref{fig:benchmarks-groupfilter}.

@@

The fourth set of queries stress the merger stage. We reuse the filter,
grouper and group filter queries that produce a $1.0$ ratio. These queries are
then run in two separate branches to produce identical filtered group records.
The merger then applies match rules to produce different output to input
ratios. The groups that are merged are again written as \texttt{flow-tools}
files using the same \texttt{zlib} compression level. The ratio of the number
of merged groups to twice\footnote{Each branch pushes the entire trace as an
input to the merger.} the number of flow records in the original trace is
plotted against time. The results are shown in Fig.
\ref{fig:benchmarks-merger}. A data point for SiLK for the $0.2$ ratio is not
available since the \ac{NFQL} query executed at that data point uses
\texttt{OR} expressions, which are not supported by SiLK.

@@

It can be seen that the merger is the most performance critical stage of the
\ac{NFQL} pipeline thus far. It is due to the fact that the merger is working
on twice the number of flow records than any other previous stage. In
addition, each branch is writing the results of the filter, grouper and group
filter stage to \texttt{flow-tools} files. As a result, the amount of disk I/O
involved is twice as much as well. Even though each branch is delegated to a
separate core using affinity masks, most of time is taken in writing these
results to the file. These results although look less promising, they are way
better than the original \ac{NFQL} merger implementation. The optimized
merger implementation takes advantage of sorted nature of filtered groups and
therefore can significantly reduce the number of merger matches. It also
writes a merged group record to a file only once despite the number of times
it has matched. Without these optimizations, running such queries on the
merger kept the CPU churning for days without results.

The last set of queries stress the ungrouper stage. They reuse the entire
merger queries as is, but enable the ungrouper as well. This means, that the
ungrouper now attempts to unfold the merged groups returned by the merger to
write their member flow records to \texttt{flow-tools} files.  However, since
the merger receives each flow record as its own filtered group, each merged
group has only one member. As a result, the ungrouper ends up rewriting the
merged groups as \texttt{flow-tools} files using the same \texttt{zlib}
compression level. This means that the execution engine ends up taking twice
the amount of time than the merger. It is important to note, that SiLK does
not have such an equivalent ungrouping tool and is therefore not considered in
this final evaluation.

### 2013-01-15 sections/implementation/execution-workflow.tex
re-read the paper; few grammatical corrections

A custom C library has been written to directly read/write data stored in
\texttt{flow-tools} format. The library sequentially reads the flow-records
into memory to support random access required for relative filtering. Each
flow-record is stored in a \texttt{char} array and the offsets to each field
are stored in separate \texttt{structs} as shown in listing
\ref{lst:c-lib-struct}. The array of such records are indexed allowing fast
retrieval in $O(1)$ time.

@@

In order to be able to make comparisons on the field offsets of a term, the
comparator needs to know the type of the comparison and the length of the
field offset. This information is parsed by execution engine once the query is
read and is therefore not available at compile time.  In order to subvert the
need to define complex branching statements, a dedicated comparator is defined
for every possible field length and comparison operation. A Python script
generates C source code for these comparators at compile time conforming to
the structure shown in listing \ref{lst:filter-struct}. This allows the term
definitions to make runtime calls using a function name derived from the
combination of operation type and field length.

@@

The execution engine, as defined by the flow query language must read all the
flow records of a supplied trace into memory before starting the processing
pipeline.  Since, the filter stage uses the supplied set of absolute rules to
make a decision on whether or not to keep a flow record; it has to pass
through the whole in-memory recordset \emph{again} to produce the filter
results. This technique involves multiple linear runs on the trace and
therefore slows down when the ratio of number of filtered records to the total
number of flow-records is high. We optimized this behavior in \texttt{nfql} by
merging the filter stage with in-memory read of the trace. This means, a
decision on whether or not to make room for a record in memory and eventually
hold a pointer for it in filter results is done upfront as soon as the record
is read from the trace. In addition, if a request to write the filter stage
results to a flow-tools file has been made, the writes are also made as soon
the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. We used the publicly available Trace 7 from the
SimpleWeb \cite{simpleweb} to compare the performance of \texttt{nfql} against
the one defined by the flow query language as shown in Fig.
\ref{fig:fv1-fv2-filter}. The filter stage implementation with these
optimizations runs 10 times faster and is more pronounced on higher ratios.

@@

The merger as formulated in the flow query language needs to match each group
record from one branch with every other record of each branch. This leads to a
complexity of $O(g^m)$ where $g$ is the number of filtered group records and
$m$ is the number of branches. The possible number of tries when matching
group records, however, can be reduced by sorting the group records on the
field offsets used for a match. \texttt{nfql} optimizes the merger to skip
over iterator permutations when a state of a current field offset value may
not allow any further match beyond the index in the current branch.  For such
an optimization to work, the filtered group records must be sorted in the
order of field offsets specified in the merger clause.  Specifying the
filtered group records in any other order may lead to undefined behavior. This
means, that if the same field offsets were used in the grouper stage, the
terms in the group clause can be rearranged by the query designer to align
with the order of terms in the merger clause.

>>>>

A custom C library has been written to directly read/write data stored in
\texttt{flow-tools} format. The library sequentially reads the flow-records
into memory to support random access required for relative filtering. Each
flow-record is stored in a \texttt{char} array and the offsets to each field
are stored in separate \texttt{structs} as shown in Listing
\ref{lst:c-lib-struct}. The array of such records are indexed allowing fast
retrieval in $O(1)$ time.

@@

In order to be able to make comparisons on the field offsets of a term, the
comparator needs to know the type of the comparison and the length of the
field offset. This information is parsed by execution engine once the query is
read and is therefore not available at compile time.  In order to subvert the
need to define complex branching statements, a dedicated comparator is defined
for every possible field length and comparison operation. A Python script
generates C source code for these comparators at compile time conforming to
the structure shown in Listing \ref{lst:filter-struct}. This allows the term
definitions to make runtime calls using a function name derived from the
combination of operation type and field length.

@@

The execution engine, as defined by the flow query language must read all the
flow records of a supplied trace into memory before starting the processing
pipeline.  Since, the filter stage uses the supplied set of absolute rules to
make a decision on whether or not to keep a flow record; it has to pass
through the whole in-memory recordset \emph{again} to produce the filter
results. This technique involves multiple linear runs on the trace and
therefore slows down when the ratio of number of filtered records to the total
number of flow-records is high. We optimized this behavior in \texttt{nfql} by
merging the filter stage with in-memory read of the trace. This means, a
decision on whether or not to make room for a record in memory and eventually
hold a pointer for it in filter results is done upfront as soon as the record
is read from the trace. In addition, if a request to write the filter stage
results to a \texttt{flow-tools} file has been made, the writes are also made as soon
the filter stage decision is available, thereby allowing
reading-filtering-writing to happen in $O(n)$ time, where $n$ is the number of
records in the trace. We used the publicly available Trace 7 from the
SimpleWeb \cite{simpleweb} to compare the performance of \texttt{nfql} against
the one defined by the flow query language as shown in Fig.
\ref{fig:fv1-fv2-filter}. The filter stage implementation with these
optimizations runs 10 times faster and is more pronounced on higher ratios.

@@

The merger as formulated in the flow query language needs to match each group
record from one branch with every other record of each branch. This leads to a
complexity of $O(g^m)$ where $g$ is the number of filtered group records and
$m$ is the number of branches. The possible number of tries when matching
group records, however, can be reduced by sorting the group records on the
field offsets used for a match. \texttt{nfql} optimizes the merger to skip
over iterator permutations when a state of a current field offset value may
not allow any further match beyond the index in the current branch.  For such
an optimization to work, the filtered group records must be sorted in the
order of field offsets specified in the merger clause.  Specifying the
filtered group records in any other order may lead to undefined behavior. This
means, that if the same field offsets were used in the grouper stage, the
terms in the grouper clause can be rearranged by the query designer to align
with the order of terms in the merger clause.
